Freezing of a Weeks-Chandler-Anderson Fluid using classical Density Functional Theory

### 2020-11-25 papers/thesis-kirstie/thesis.tex
cleanup
fixed spelling and removed white borders around plots

\abstract{
  We use Classical Density Functional Theory (cDFT) to predict the 
  freezing of a Weeks-Chandler-Anderson (WCA) fluid. A WCA fluid 
  consists of soft spheres that exhibit the repulsive portion of the 
  Leonard-Jones potential widely used to model the interaction of two 
  neutral atoms. Our cDFT functional is constructed using Soft Fundamental 
  Theory (SFMT) with Tarazona's tensor expression for $\Phi_3$, and Schmidt's
  erf model in which we have incorporated temperature dependent
  parameters to fit the results to those predicted by Monte-Carlo simulations.
  Our functional was successful in predicting the freezing of the WCA fluid 
  with a fairly good match to Monte-Carlo simulations. This result suggests 
  that SFMT may provide a solid foundation for future density functionals 
  that will model real fluids.
}

@@

Soft Fundamental Measure Theory (SFMT) will be used to form the 
functional for the excess Helmholtz free energy. 
SFMT is a method that makes the task of 
forming an expression for the excess Helmholtz free energy of a particular 
model fluid much easier since the interaction that each sphere, or atom, 
has with every other sphere does not have to be taken explicitly into account.
Once the functional for the Helmholtz free energy is formed, classical 
Density Functional Theory is used to find the equilibrium 
Helmholtz free energy at a particular temperature and density. 
According to cDFT the equilbrium 
Helmholtz free energy of the system can be found by varying the number 
density profile $n(\vec{r})$ until the free energy is minimized. 
Results will be compared to those from Monte-Carlo simulations.

@@

Real fluids and their behavior are very complex from a physicist's point 
of view. Although a water molecule, for example, is composed of only three
atoms, the ever-changing redistribution of charges within the molecules, 
and their interaction with charges from other molecules make it difficult 
to study a fluid composed of water molecules directly. The task becomes 
even more daunting considering the complicated molecular motion of each 
water molecule, and the sheer number of molecules that must be taken 
into account. 
The problem of understanding the behavoir of real liquids is better addressed 
by using models which reduce a complex picture to a simple one. Modifications, 
or perturbations, can then be incorporated into the model making it 
progressively more complex until it mimics the behavior of a real 
liquid as closely as possible.

@@

There are many interatomic potentials which can be used to model a fluid 
each of which describes the interaction between two real atoms to some 
degree of accuracy. The simplest, and oldest, model used to describe a 
fluid is the hard-sphere model. The hard-sphere model envisions a fluid 
to be composed of hard spheres like, for example, a collection of marbles. 
This model came about by considering the near incompressibilty of liquids 
indicating that the particles making up the liquid strongly resist being 
pushed closer together. This behavior can be described by an interatomic 
potential energy that is zero when two spheres are not in contact, but 
goes instantly to infinity when contact is made. The inablilty to push 
real atoms closer together is due to the repulsion of the electrons in 
each atom, and is the dominating interparticle electrostatic interaction. 
Thus, the hard-sphere model - which is the simplest, but still effective 
representation of this replusive behavior that dominates the interaction 
between real atoms, is currently the most widely used model for fluids.

@@

Another fairly simple, but relatively accurate interatomic potential widely 
used to represent generic, neutral atoms is the Leonard-Jones potential. 
The Leonard-Jones potential describes both the long-range attraction, 
and the short-range replusion that occurs between two nuetral atoms. 
The attraction between two nuetral atoms is due to fleeting dipoles that 
form spontaneously when electrons are redistributed as atoms come into 
close proximty. These are often characterized as Van der Walls, or London 
dispersion forces. The attraction grows stronger as two atoms move closer 
together until their atomic orbitals begin to overlap. 
At this point the electrons in each atom begin to oppose the overlap of 
the two atoms through their replusion. The repulsion increases rapidly and 
approaches infinity as the distance between the atoms conintues to decrease.

\begin{figure}[h!]
    \centering
    \includegraphics[height=7cm]{LJ_Potential.pdf}
    \caption{Leonard-Jones Potential for $\sigma=1$, $\epsilon=1$}
    \label{fig:LJ_potential}
  \end{figure}

The combined attractive and repulsive effects produce the overall shape of 
the Leonard-Jones interatomic potential shown in Figure \ref{fig:LJ_potential}. 
The potential is zero when the distance r between the centers of two atoms 
is equal to $\sigma$ (shown in the figure at r/$\sigma$=1). At this point 
the attraction bewteen the atoms just cancels their repulsion. As the distance 
between the centers of the two atoms decreases below $\sigma$, the potential 
rapidly approaches positive infinity corresponding to infinite repulsion. 
Above $\sigma$ the potential becomes negative indicating a net attraction 
between the atoms. The net attraction is maximized at a distance of 
r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ where R is the radius of 
the sphere given by $R={2^{-\frac{5}{6}}}\sigma$. At this point the depth of
the potential well, given by epsilon $\epsilon$, is a maximum. As r increases 
further, the net attraction decreases and the potential approaches zero
asymptotically. The attractive and repulsive effects can be separated into two
potentials, shown with different colors in Figure \ref{fig:LJ_potential_2parts}, 
which when added together regenerate the full Leonard-Jones potential.

\begin{figure}[h!]
    \centering
    \includegraphics[height=7cm]{LJ_Potential_2part.pdf}
    \caption{Leanard-Jones Potential broken into two parts for $\sigma=1$, $\epsilon=1$}
    \label{fig:LJ_potential_2parts}
  \end{figure}

@@

The repulsive portion of the Leonard-Jones potential forms the potential for a 
Weeks-Chandler-Anderson (WCA) model fluid~\cite{andersen1971relationship}, 
shown in Figure \ref{fig:WCA_potential}. 
The WCA potential is described by the equation: 
\begin{align} \label{eq:VWCA}
    V_{WCA}=\left\{\begin{array}{rcl} {4\epsilon{\left[\left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right]}+\epsilon} & \mbox{for} & 0<r<{2R} \\ 0 & \mbox{for} & r>2R \end{array}\right.
\end{align} 
As seen in Figure \ref{fig:WCA_potential} the WCA potential decreases 
from infinity as the distance between the centers of the atoms r increases, 
and becomes, and remains, essentially zero when 
r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ 
where R is the radius of the sphere.

As with the Lenard-Jones potential, the repulsive WCA potential increases 
rapidly and approches infinity as the two spheres begin to overlap. This 
is a more realistic repulsive potential than that of the Hard Sphere model 
which goes instantly to infinity when the two spheres make contact (at r=2R), 
and is zero otherwise. Because the WCA spheres have no hard boundary in 
their interatomic potential, they are called ``soft spheres'' and their 
gradually fading potential makes them slightly ``squishy'', as illustrated in
Figure~\ref{fig:TwoSpheres}.

@@

%\newpage
\section{The Theory of Freezing and Maxwell's Double Tangent Construction}

@@

For pure substances, a phase transition occurs along a line on a 
Pressure-Temperature (P-T) phase diagram, such as can be seen in the phase 
diagram shown in Figure~\ref{fig:P-T_Diagram}. This phase diagram is 
representative of some arbitrary, unknown substance composed of atoms, 
or molecules which exibit both attraction and repulsion. However, since 
the atoms in a WCA fliud do not exhibit attraction there will not be a 
dividing line between the gas and liquid phases, and so the term "liquid" 
or "gas" is not used, but simply "fluid". Points along the line that 
separates the solid and the fluid phases indicate a phase transition, 
and show that the solid and the fluid both have the same temperature and 
the same pressure at a phase transistion.

@@

Two such curves are shown in Figure~\ref{fig:MaxwellDT}, one for the liquid 
and one for the solid at a fixed temperature. The slope of the curves is 
related to the pressure. Since the pressure of the 
liquid is the same as the pressure of the solid at the point of transistion, 
the density of the liquid and the density of the solid at the point of 
freezing can be found from identifying the inverse densties at which the 
tangents to each curve have the same slope. This process is called 
Maxwell's Double Tangent Construction.

\begin{figure}[h!]
    \centering
     \includegraphics[height=8cm]{figs/MaxwellDTC-Fig1.pdf}
    \caption{Identifying densities at the solid-liquid transition using 
    Maxwell's Double Tangent Construction}
    \label{fig:MaxwellDT}
  \end{figure}

@@

The pressure at the transistion point can be obtained from the point of 
intersection that occurs when the Gibbs free energy per atom is plotted 
against the pressure for both the liquid and the solid at fixed temperature, 
as shown in Figure~\ref{fig:GibbsvsP}. These curves are constructed from 
the slope and cooresponding y-intercept data collected from lines tangent 
along the curves on the the Helmholtz free energy per atom verses inverse 
density plots (like that of Figure~\ref{fig:MaxwellDT}) for both the 
liquid and the solid at a given temperature.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[height=6cm]{T-n_Diagram.png}
    \caption{Conceptual Temperature-Density Phase Diagram used to identify 
    the number densities of the liquid and the solid during a phase 
    transistion. The gray region is where the liquid and the solid coexist, 
    and are therefore not pure states.}
    \label{fig:T-n_Diagram}
  \end{figure}

@@

A description of cDFT often begins with considering the Grand Free 
Energy defined as
\begin{equation}\Phi=F-\mu{N}\end{equation}
which can then be expressed in terms of a varrying number density 
profile $n(\vec r)$
\begin{equation}\label{GrandFE}\Phi= F_{int} +\int n(\vec{r})\phi{(\vec r)}d\vec{r}-\mu\int n(\vec r)d\vec{r}\end{equation}
where $\phi=\frac{V_{ext}}{N}$ 
is the externally 
applied potential energy $V_{ext}(\vec r)$ per particle, 
and the Helmholtz free energy is expressed as
\begin{equation}F = F_{int} + \int n(\vec{r})\phi{(\vec r)}d\vec{r}\end{equation}

@@

The minimization can be illustrated by considering that according to the 
second law of thermodynamics, a system that is not in equilibrium will 
exprience spontanteous processes that increase the total entropy of a 
system and its surroundings until equilibrium is reached, 
$\Delta{S}_{System + Surroundings} \geq 0$. Here ``surroundings" means 
the entire universe minus the system. This, together with the conservation 
of energy (the first law of thermodynamics) which says that energy lost 
(or gained) by the system is gained (or lost) by the surroundings 
$\Delta{U}_{System}=-\Delta{U}_{Surroundings}$, gives the result that 
changes in the free energy will be negative for spontaneous processes 
when the natural variables of the free energy are held fixed~\cite{schroeder}. 
A general proof of minimization 
can be found in Theory of Simple Fluids by Hansen and McDonald~\cite{Hansen}.

@@

\begin{equation}{\label{Fidealmultipleint}}{=-k_BT\ln\left(V^N\frac{1}{N!}\frac{1}{h^{3}}\int{d\vec{p}_1}~e^\frac{-\frac{|\vec{p}_1|^2}{2m}}{k_BT}\frac{1}{h^{3}}\int{d\vec{p}_2}~e^\frac{-\frac{|\vec{p}_2|^2}{2m}}{k_BT}{...}\frac{1}{h^{3}}\int{d\vec{p}_N}~e^\frac{-\frac{|\vec{p}_N|^2}{2m}}{k_BT}\right)}\end{equation}
These integrals can conveniently be written in terms of the DeBroglie 
wavelength $\Lambda$, or the quantum volume $\text{n}_\text{Q}$
\begin{equation}{\frac{1}{h^{3}}\int{d\vec{p}}~e^\frac{-\frac{|\vec{p}|^2}{2m}}{k_BT}=\left(\sqrt{\frac{k_BTm}{2\pi\hbar^2}}\right)^3=\frac{1}{\Lambda^{3}}}=\text{n}_\text{Q}\end{equation} 
where \begin{equation}{\Lambda =\sqrt{\frac{2\pi\hbar^2}{k_BTm}}}\end{equation} 
Rewriting Equation~\ref{Fidealmultipleint} in terms of $\Lambda$ gives
\begin{equation}{F_{ideal}= -k_BT[\ln(V^N)-\ln(N!) - \ln(\Lambda^{3N})]}\end{equation}
Using Stirling's approximation \begin{equation}{\ln(N!)=N\ln{N}-N}\end{equation} 
this becomes
\begin{equation}{F_{ideal}= k_BT[N\ln(\frac{N\Lambda^{3}}{V}{)-N}]}\end{equation} 
Rewriting the equation in terms of the number density $n=\frac{N}{V}$, 
and the quantum volume, $\text{n}_\text{Q}$ gives
\begin{equation}{F_{ideal}= k_BT\left[\frac{N}{V}\ln{\left(\frac{N}{V}\frac{1}{\text{n}_\text{Q}}\right)}-\frac{N}{V}\right]}V\end{equation}
\begin{equation}\label{eq:Fideal}{F_{ideal}= k_BT[n\ln(n/\text{n}_\text{Q})-n]V}\end{equation}   
For a spatially varying number density, the ideal free energy in 
Equation~\ref{eq:Fideal} becomes
\begin{equation}\label{eq:Fideal-n(r)}{F_{ideal}[n(\vec{r})]= k_BT\int_{Vol}[n(\vec{r})\ln(n(\vec{r})/\text{n}_\text{Q})-n(\vec{r})]d\vec{r}}\end{equation}

@@

  \begin{figure}[h!]
    \centering
    \includegraphics[height=7cm]{figs/diagrammic.pdf}
    \caption{Two-pair Interactions $f_{ij}f_{kl}$ for N=4 atoms. Interations 
    between individual pairs of atoms are shown in green. Interactions 
    between pairs of atoms with an atom shared in common are shown in red. 
    At low densities only the interactions shown in green are considered, 
    since atoms are not likely to be clustered.}
    \label{fig:diagrammic}
  \end{figure}

At low densities, it is not likely to find two or more atoms in close 
proximity, and so only interactions between individual pairs of spheres, 
such as those shown in green in Figure~\ref{fig:diagrammic}, are considered.
Each pair can then be described by a function of the distance between the 
two atoms in the pair. 
With a careful counting of the resulting number of such pairs, the excess 
free energy can then be written~\cite{schroeder} as
\begin{equation}\label{eq:Fexcess}{\frac{F_{excess}}{N}=k_BTn\left(-\frac{1}{2}\int{f(r)}{d\vec{r}}\right)}\end{equation} 
in the thermodynamic limit of large N 
where the quantity in parenthesis can be recognized 
as the second virial expanision coefficient typically called $B_2$. 
It is important that the expression for the excess Helmholtz free energy 
correctly reduces to this result in the low-density limit. 
Outside of the low density limit, it is quite difficult to evaluate the 
excess Helmholtz free energy given in Equation~\ref{eq:Fexcess-simplified}.

Once an expression for the Helmholtz free energy as a functional of the
number density profile $n(\vec r)$ is formed then, according to classical 
Density Functional Theory, the equilibrium value of the 
Helmholtz energy, and its cooresonding equilibrium density profile can be
found by varying the number density profile
through all possible configurations at some fixed temperature
until the free energy is minimized.  
In this section 
a functional for the Helmholtz free energy was formed by finding 
general expressions for the ideal and the excess portions of the free 
energy, and adding them together.
\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation}
The functional for the ideal portion of the Helmholtz free energy is given by 
Equation~\ref{eq:Fideal-n(r)}, and a functional for the excess portion of the Helmholtz 
free energy is given in Equation~\ref{eq:Fexcess-simplified}. The specific 
form for the excess Helmholtz free energy depends on the particular 
fluid being modeled as defined by the  
interatomic potential energy V(r) between the spheres, which can make
forming an expression difficult. In this paper, Soft Fundamental Measure
Theory will be used to form an epxression for the excess Helmholtz free 
energy, as explained in the next section.

\section{Soft Fundamental Measure Theory (SFMT)}
In order to apply classical Density Functional Theory to a system that has
a fixed temperature, volume, and nubmer of particles, a functional for 
the Helmholtz free energy must first be formed. An expression for the ideal 
portion of the Helmnoltz free energy is given by Equation~\ref{eq:Fideal-n(r)}, 
but forming an expression for the excess Helmholtz free
energy from Equation~\ref{eq:Fexcess-simplified} for the particular fluid 
being modeled can be challenging. 
Soft Fundamental Measure Theory (SFMT) is a method for forming an 
expression for the excess Helmholtz free energy of a Soft-sphere fluid
that does not depend on  explicitly taking into account the interaction 
that each sphere with every other sphere in the fluid as required by 
Equation~\ref{eq:Fexcess-simplified}. In the next chapter on methods, we
will use an SFMT functional that we adapt to fit the soft-sphere WCA fluid.

>>>>

\abstract{
  We use Classical Density Functional Theory (cDFT) to predict the 
  freezing of a Weeks-Chandler-Anderson (WCA) fluid. A WCA fluid 
  consists of soft spheres that exhibit the repulsive portion of the 
  Lennard-Jones potential widely used to model the interaction of two 
  neutral atoms. Our cDFT functional is constructed using Soft Fundamental 
  Theory (SFMT) with Tarazona's tensor expression for $\Phi_3$, and Schmidt's
  erf model in which we have incorporated temperature dependent
  parameters to fit the results to those predicted by Monte-Carlo simulations.
  Our functional was successful in predicting the freezing of the WCA fluid 
  with a fairly good match to Monte-Carlo simulations. This result suggests 
  that SFMT may provide a solid foundation for future density functionals 
  that will model real fluids.
}

@@

Soft Fundamental Measure Theory (SFMT) will be used to form the 
functional for the excess Helmholtz free energy. 
SFMT is a method that makes the task of 
forming an expression for the excess Helmholtz free energy of a particular 
model fluid much easier since the interaction that each sphere, or atom, 
has with every other sphere does not have to be taken explicitly into account.
Once the functional for the Helmholtz free energy is formed, classical 
Density Functional Theory is used to find the equilibrium 
Helmholtz free energy at a particular temperature and density. 
According to cDFT the equilibrium 
Helmholtz free energy of the system can be found by varying the number 
density profile $n(\vec{r})$ until the free energy is minimized. 
Results will be compared to those from Monte-Carlo simulations.

@@

Real fluids and their behavior are very complex from a physicist's point 
of view. Although a water molecule, for example, is composed of only three
atoms, the ever-changing redistribution of charges within the molecules, 
and their interaction with charges from other molecules make it difficult 
to study a fluid composed of water molecules directly. The task becomes 
even more daunting considering the complicated molecular motion of each 
water molecule, and the sheer number of molecules that must be taken 
into account. 
The problem of understanding the behavior of real liquids is better addressed 
by using models which reduce a complex picture to a simple one. Modifications, 
or perturbations, can then be incorporated into the model making it 
progressively more complex until it mimics the behavior of a real 
liquid as closely as possible.

@@

There are many interatomic potentials which can be used to model a fluid 
each of which describes the interaction between two real atoms to some 
degree of accuracy. The simplest, and oldest, model used to describe a 
fluid is the hard-sphere model. The hard-sphere model envisions a fluid 
to be composed of hard spheres like, for example, a collection of marbles. 
This model came about by considering the near incompressibility of liquids 
indicating that the particles making up the liquid strongly resist being 
pushed closer together. This behavior can be described by an interatomic 
potential energy that is zero when two spheres are not in contact, but 
goes instantly to infinity when contact is made. The inability to push 
real atoms closer together is due to the repulsion of the electrons in 
each atom, and is the dominating interparticle electrostatic interaction. 
Thus, the hard-sphere model - which is the simplest, but still effective 
representation of this repulsive behavior that dominates the interaction 
between real atoms, is currently the most widely used model for fluids.

@@

Another fairly simple, but relatively accurate interatomic potential widely 
used to represent generic, neutral atoms is the Lennard-Jones potential. 
The Lennard-Jones potential describes both the long-range attraction, 
and the short-range repulsion that occurs between two neutral atoms. 
The attraction between two neutral atoms is due to fleeting dipoles that 
form spontaneously when electrons are redistributed as atoms come into 
close proximity. These are often characterized as Van der Walls, or London 
dispersion forces. The attraction grows stronger as two atoms move closer 
together until their atomic orbitals begin to overlap. 
At this point the electrons in each atom begin to oppose the overlap of 
the two atoms through their repulsion. The repulsion increases rapidly and 
approaches infinity as the distance between the atoms continues to decrease.

\begin{figure}[h!]
    \centering
    \includegraphics[height=7cm]{LJ_Potential.pdf}
    \caption{Lennard-Jones Potential for $\sigma=1$, $\epsilon=1$}
    \label{fig:LJ_potential}
  \end{figure}

The combined attractive and repulsive effects produce the overall shape of 
the Lennard-Jones interatomic potential shown in Figure \ref{fig:LJ_potential}. 
The potential is zero when the distance r between the centers of two atoms 
is equal to $\sigma$ (shown in the figure at r/$\sigma$=1). At this point 
the attraction between the atoms just cancels their repulsion. As the distance 
between the centers of the two atoms decreases below $\sigma$, the potential 
rapidly approaches positive infinity corresponding to infinite repulsion. 
Above $\sigma$ the potential becomes negative indicating a net attraction 
between the atoms. The net attraction is maximized at a distance of 
r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ where R is the radius of 
the sphere given by $R={2^{-\frac{5}{6}}}\sigma$. At this point the depth of
the potential well, given by epsilon $\epsilon$, is a maximum. As r increases 
further, the net attraction decreases and the potential approaches zero
asymptotically. The attractive and repulsive effects can be separated into two
potentials, shown with different colors in Figure \ref{fig:LJ_potential_2parts}, 
which when added together regenerate the full Lennard-Jones potential.

\begin{figure}[h!]
    \centering
    \includegraphics[height=7cm]{LJ_Potential_2part.pdf}
    \caption{Lennard-Jones Potential broken into two parts for $\sigma=1$, $\epsilon=1$}
    \label{fig:LJ_potential_2parts}
  \end{figure}

@@

The repulsive portion of the Lennard-Jones potential forms the potential for a 
Weeks-Chandler-Anderson (WCA) model fluid~\cite{andersen1971relationship}, 
shown in Figure \ref{fig:WCA_potential}. 
The WCA potential is described by the equation: 
\begin{align} \label{eq:VWCA}
    V_{WCA}=\left\{\begin{array}{rcl} {4\epsilon{\left[\left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right]}+\epsilon} & \mbox{for} & 0<r<{2R} \\ 0 & \mbox{for} & r>2R \end{array}\right.
\end{align} 
As seen in Figure \ref{fig:WCA_potential} the WCA potential decreases 
from infinity as the distance between the centers of the atoms r increases, 
and becomes, and remains, essentially zero when 
r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ 
where R is the radius of the sphere.

As with the Lenard-Jones potential, the repulsive WCA potential increases 
rapidly and approaches infinity as the two spheres begin to overlap. This 
is a more realistic repulsive potential than that of the Hard Sphere model 
which goes instantly to infinity when the two spheres make contact (at r=2R), 
and is zero otherwise. Because the WCA spheres have no hard boundary in 
their interatomic potential, they are called ``soft spheres'' and their 
gradually fading potential makes them slightly ``squishy'', as illustrated in
Figure~\ref{fig:TwoSpheres}.

@@

\section{The Theory of Freezing and Maxwell's Double Tangent Construction}

@@

For pure substances, a phase transition occurs along a line on a 
Pressure-Temperature (P-T) phase diagram, such as can be seen in the phase 
diagram shown in Figure~\ref{fig:P-T_Diagram}. This phase diagram is 
representative of some arbitrary, unknown substance composed of atoms, 
or molecules which exhibit both attraction and repulsion. However, since 
the atoms in a WCA fluid do not exhibit attraction there will not be a 
dividing line between the gas and liquid phases, and so the term "liquid" 
or "gas" is not used, but simply "fluid". Points along the line that 
separates the solid and the fluid phases indicate a phase transition, 
and show that the solid and the fluid both have the same temperature and 
the same pressure at a phase transition.

@@

Two such curves are shown in Figure~\ref{fig:MaxwellDT}, one for the liquid 
and one for the solid at a fixed temperature. The slope of the curves is 
related to the pressure. Since the pressure of the 
liquid is the same as the pressure of the solid at the point of transition, 
the density of the liquid and the density of the solid at the point of 
freezing can be found from identifying the inverse densities at which the 
tangents to each curve have the same slope. This process is called 
Maxwell's Double Tangent Construction.

\begin{figure}[h!]
    \centering
     \includegraphics[height=8cm]{figs/MaxwellDTC-Fig1.pdf}
    \caption{Densities at the solid-liquid transition are identified using 
    Maxwell's Double Tangent Construction}
    \label{fig:MaxwellDT}
  \end{figure}

@@

The pressure at the transition point can be obtained from the point of 
intersection that occurs when the Gibbs free energy per atom is plotted 
against the pressure for both the liquid and the solid at fixed temperature, 
as shown in Figure~\ref{fig:GibbsvsP}. These curves are constructed from 
the slope and corresponding y-intercept data collected from lines tangent 
along the curves on the the Helmholtz free energy per atom verses inverse 
density plots (like that of Figure~\ref{fig:MaxwellDT}) for both the 
liquid and the solid at a given temperature.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[height=6cm]{T-n_Diagram.png}
    \caption{Conceptual Temperature-Density Phase Diagram used to identify 
    the number densities of the liquid and the solid during a phase 
    transition. The gray region is where the liquid and the solid coexist, 
    and are therefore not pure states.}
    \label{fig:T-n_Diagram}
  \end{figure}

@@

A description of cDFT often begins with considering the Grand Free 
Energy defined as
\begin{equation}\Phi=F-\mu{N}\end{equation}
which can then be expressed in terms of a varying number density 
profile $n(\vec r)$
\begin{equation}\label{GrandFE}\Phi= F_{int} +\int n(\vec{r})\phi{(\vec r)}d\vec{r}-\mu\int n(\vec r)d\vec{r}\end{equation}
where $\phi=\frac{V_{ext}}{N}$ 
is the externally 
applied potential energy $V_{ext}(\vec r)$ per particle, 
and the Helmholtz free energy is expressed as
\begin{equation}F = F_{int} + \int n(\vec{r})\phi{(\vec r)}d\vec{r}\end{equation}

@@

The minimization can be illustrated by considering that according to the 
second law of thermodynamics, a system that is not in equilibrium will 
experience spontaneous processes that increase the total entropy of a 
system and its surroundings until equilibrium is reached, 
$\Delta{S}_{System + Surroundings} \geq 0$. Here ``surroundings" means 
the entire universe minus the system. This, together with the conservation 
of energy (the first law of thermodynamics) which says that energy lost 
(or gained) by the system is gained (or lost) by the surroundings 
$\Delta{U}_{System}=-\Delta{U}_{Surroundings}$, gives the result that 
changes in the free energy will be negative for spontaneous processes 
when the natural variables of the free energy are held fixed~\cite{schroeder}. 
A general proof of minimization 
can be found in Theory of Simple Fluids by Hansen and McDonald~\cite{Hansen}.

@@

\begin{equation}{\label{Fidealmultipleint}}{=-k_BT\ln\left(V^N\frac{1}{N!}\frac{1}{h^{3}}\int{d\vec{p}_1}~e^\frac{-\frac{|\vec{p}_1|^2}{2m}}{k_BT}\frac{1}{h^{3}}\int{d\vec{p}_2}~e^\frac{-\frac{|\vec{p}_2|^2}{2m}}{k_BT}{...}\frac{1}{h^{3}}\int{d\vec{p}_N}~e^\frac{-\frac{|\vec{p}_N|^2}{2m}}{k_BT}\right)}\end{equation}
These integrals can conveniently be written in terms of the de Broglie 
wavelength $\Lambda$, or the quantum volume $\text{n}_\text{Q}$
\begin{equation}{\frac{1}{h^{3}}\int{d\vec{p}}~e^\frac{-\frac{|\vec{p}|^2}{2m}}{k_BT}=\left(\sqrt{\frac{k_BTm}{2\pi\hbar^2}}\right)^3=\frac{1}{\Lambda^{3}}}=\text{n}_\text{Q}\end{equation} 
where \begin{equation}{\Lambda =\sqrt{\frac{2\pi\hbar^2}{k_BTm}}}\end{equation} 
Rewriting Equation~\ref{Fidealmultipleint} in terms of $\Lambda$ gives
\begin{equation}{F_{ideal}= -k_BT[\ln(V^N)-\ln(N!) - \ln(\Lambda^{3N})]}\end{equation}
Using Stirling's approximation \begin{equation}{\ln(N!)=N\ln{N}-N}\end{equation} 
this becomes
\begin{equation}{F_{ideal}= k_BT[N\ln(\frac{N\Lambda^{3}}{V}{)-N}]}\end{equation} 
Rewriting the equation in terms of the number density $n=\frac{N}{V}$, 
and the quantum volume, $\text{n}_\text{Q}$ gives
\begin{equation}{F_{ideal}= k_BT\left[\frac{N}{V}\ln{\left(\frac{N}{V}\frac{1}{\text{n}_\text{Q}}\right)}-\frac{N}{V}\right]}V\end{equation}
\begin{equation}\label{eq:Fideal}{F_{ideal}= k_BT[n\ln(n/\text{n}_\text{Q})-n]V}\end{equation}   
For a spatially varying number density, the ideal free energy in 
Equation~\ref{eq:Fideal} becomes
\begin{equation}\label{eq:Fideal-n(r)}{F_{ideal}[n(\vec{r})]= k_BT\int_{Vol}[n(\vec{r})\ln(n(\vec{r})/\text{n}_\text{Q})-n(\vec{r})]d\vec{r}}\end{equation}

@@

  \begin{figure}[h!]
    \centering
    \includegraphics[height=7cm]{figs/diagrammic.pdf}
    \caption{Two-pair Interactions $f_{ij}f_{kl}$ for N=4 atoms. Interactions 
    between individual pairs of atoms are shown in green. Interactions 
    between pairs of atoms with an atom shared in common are shown in red. 
    At low densities only the interactions shown in green are considered, 
    since atoms are not likely to be clustered.}
    \label{fig:diagrammic}
  \end{figure}

At low densities, it is not likely to find two or more atoms in close 
proximity, and so only interactions between individual pairs of spheres, 
such as those shown in green in Figure~\ref{fig:diagrammic}, are considered.
Each pair can then be described by a function of the distance between the 
two atoms in the pair. 
With a careful counting of the resulting number of such pairs, the excess 
free energy can then be written~\cite{schroeder} as
\begin{equation}\label{eq:Fexcess}{\frac{F_{excess}}{N}=k_BTn\left(-\frac{1}{2}\int{f(r)}{d\vec{r}}\right)}\end{equation} 
in the thermodynamic limit of large N 
where the quantity in parenthesis can be recognized 
as the second virial expansion coefficient typically called $B_2$. 
It is important that the expression for the excess Helmholtz free energy 
correctly reduces to this result in the low-density limit. 
Outside of the low density limit, it is quite difficult to evaluate the 
excess Helmholtz free energy given in Equation~\ref{eq:Fexcess-simplified}.

Once an expression for the Helmholtz free energy as a functional of the
number density profile $n(\vec r)$ is formed then, according to classical 
Density Functional Theory, the equilibrium value of the 
Helmholtz energy, and its corresponding equilibrium density profile can be
found by varying the number density profile
through all possible configurations at some fixed temperature
until the free energy is minimized.  
In this section 
a functional for the Helmholtz free energy was formed by finding 
general expressions for the ideal and the excess portions of the free 
energy, and adding them together.
\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation}
The functional for the ideal portion of the Helmholtz free energy is given by 
Equation~\ref{eq:Fideal-n(r)}, and a functional for the excess portion of the Helmholtz 
free energy is given in Equation~\ref{eq:Fexcess-simplified}. The specific 
form for the excess Helmholtz free energy depends on the particular 
fluid being modeled as defined by the  
interatomic potential energy V(r) between the spheres, which can make
forming an expression difficult. In this paper, Soft Fundamental Measure
Theory will be used to form an expression for the excess Helmholtz free 
energy, as explained in the next section.

\section{Soft Fundamental Measure Theory (SFMT)}
In order to apply classical Density Functional Theory to a system that has
a fixed temperature, volume, and number of particles, a functional for 
the Helmholtz free energy must first be formed. An expression for the ideal 
portion of the Helmholtz free energy is given by Equation~\ref{eq:Fideal-n(r)}, 
but forming an expression for the excess Helmholtz free
energy from Equation~\ref{eq:Fexcess-simplified} for the particular fluid 
being modeled can be challenging. 
Soft Fundamental Measure Theory (SFMT) is a method for forming an 
expression for the excess Helmholtz free energy of a Soft-sphere fluid
that does not depend on  explicitly taking into account the interaction 
that each sphere with every other sphere in the fluid as required by 
Equation~\ref{eq:Fexcess-simplified}. In the next chapter on methods, we
will use an SFMT functional that we adapt to fit the soft-sphere WCA fluid.

SFMT is an extension of Fundamental Measure Theory (FMT)
that was developed for the Hard-sphere fluid, and which can be derived from
Scaled Particle Theory (SPT).  
Scaled Particle Theory seeks a general expression for the excess chemical 
potential of a homogeneous Hard Sphere fluid which can then be integrated 
to find the excess Helmholtz free energy according to the relation
\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~}\rightarrow{~~~~~}F_{ex}=\int{\mu_{ex}dN}\end{equation}
It is an extension of the work done by Widom who showed that the excess 
chemical potential of a homogenesous fluid corresponds to the reversible 
work of inserting a sphere of radius R into a fluid consisting of hard 
spheres of radius R. 
\begin{align}
	\mu_{ex}=W_{rev}(R){~~~~}\text{work to insert a hard sphere of radius R}
\end{align}
Widom's result was obtained by considering the difference 
in the excess Helmholtz free energy when one atom is added to an otherwise 
closed system. He showed that the excess chemical potential is related to 
the ensemble average of the Boltzman factor for the potential energy 
difference $V_{diff}$ between a system with N atoms and one with N+1 atoms. 
The resulting equation, shown below, is known as Widom's Insertion 
Formula~\cite{Hansen}.
\begin{equation}\mu_{ex}=\frac{F_{ex}(N+1)-F_{ex}(N)}{\Delta{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\label{widoms-insertion-formula}{~}=-k_BT\ln\left(<e^{-\beta{V_{diff}}}>\right)\end{equation}
